---
title: "Reproducible Research: Peer Assessment 1"
output: 
  html_document:
    keep_md: true
---
###load packages
```{r}
knitr::opts_chunk$set(echo = TRUE,results = "asis")

library(timeDate)
library(plyr)
library(dplyr)
library(data.table)
library(ggplot2)
```
## Loading and preprocessing the data
*Read file*
*Add "time" column with converted "interval" column values to time format*
*Add "day" column to display weekday*
*Add "wday" logical column for weekend or not*

```{r}
data <- read.csv("activity.csv")
data$day <- weekdays(as.Date(data$date))
data$subset_key <- paste(data$day, data$interval, sep="_")
```

# Assignment Questions
## What is mean total number of steps taken per day?
### Calculate total steps, mean and median steps per day
```{r}
steps_day <- group_by(data, date)
steps_sum <- summarize(steps_day, day_sum = sum(steps, na.rm = TRUE))
steps_sum_mean <- mean(steps_sum$day_sum)
steps_sum_median <- median(steps_sum$day_sum)
```

### Histogram of total steps per day
```{r}
h<-hist(steps_sum$day_sum, breaks = 10, col = "green", main="Histogram of Total Steps Per Day", xlab = "Total Steps Per Day")
print(h)
```

```{r, echo=FALSE}
png("TotalStepsPerDay.png")
dev.set(2)
dev.copy(png, "TotalStepsPerDay.png")
dev.off()
```
### Mean of total steps per day
```{r}
steps_sum_mean
```

### Median of total steps per day
```{r}
steps_sum_median
```

## What is the average daily activity pattern?
### plot of steps taken by intervals
```{r}
steps_interval <- group_by(data, interval)
steps_by_interval <- summarize(steps_interval, interval_mean = mean(steps, na.rm = TRUE))
p<-plot(steps_by_interval$interval, steps_by_interval$interval_mean, main="Average Steps Per 5 Minute Interval", xlab="5 Minute Time Intervals", ylab="Average Steps", col = "red", type = "l")
print(p)
```

```{r, echo=FALSE}
png("DailyActivityPattern.png")
dev.set(2)
dev.copy(png, "DailyActivityPattern.png")
dev.off()
```

###Maximum average steps by interval
```{r}
step_interval_max <- steps_by_interval[which.max(steps_by_interval$interval_mean),]
step_interval_max$interval
```

## Imputing missing values
*Total number of rows with NAs*
```{r}
sum(is.na(data$steps))
```

*Create subset Data Frame for NA rows*
```{r}
data_NA <- data[is.na(data$steps),]
```

*Create subset Data Frame excluding NA rows*
```{r}
data_xNA <- data[!is.na(data$steps),]
```

*Calculate imputed value as average steps per day with intervals*
```{r}
step_impute <- group_by(data_xNA, subset_key)
step_impute_mean <- summarize(step_impute, impute_val = mean(steps, na.rm = TRUE))
```

*Substitute imputed value for NAs based on day/interval key*
```{r}
new_data <- merge(data_NA[,c("date", "interval" , "day", "subset_key")], step_impute_mean, by= "subset_key")
```

*Combine imputed xNA and NA to create new Data Frame*
```{r}
colnames(new_data)[5] <- "steps"
new_data <- new_data[c(5,2,3,4,1)]
new_data <- rbind(new_data, data_xNA)
```

*Histogram of total steps per day*
```{r}
new_steps_day <- group_by(new_data, date)
new_steps_sum <- summarize(new_steps_day, new_day_sum = sum(steps))
h<-hist(new_steps_sum$new_day_sum, breaks = 10, col = "blue", main="Histogram of Total Steps Per Day (Imputed)", xlab = "Total Steps Per Day (Imputed)")
```

```{r, echo=FALSE}
png("NewTotalStepsPerDay.png")
dev.set(2)
dev.copy(png, "NewTotalStepsPerDay.png")
dev.off()
```

*The mean and median total steps per day*
_Mean_
```{r}
new_steps_sum_mean <- mean(new_steps_sum$new_day_sum)
new_steps_sum_mean
```
_Median_
```{r}
new_steps_sum_median <- median(new_steps_sum$new_day_sum)
new_steps_sum_median
```

*Differences from before adn after imputing values mean and median increased*
_Mean steps per day difference_
```{r}
new_steps_sum_mean - steps_sum_mean
```
_Median steps per day difference_
```{r}
new_steps_sum_median - steps_sum_median
```

## Are there differences in activity patterns between weekdays and weekends?
```{r}
new_data$DayoftheWeek <- ifelse(weekdays(as.Date(new_data$date)) %in% c("Monday","Tuesday","Wednesday", "Thursday", "Friday"), "Weekday", "Weekend")
newMeanPerInterval <- new_data %>% na.omit() %>% group_by(interval, DayoftheWeek) %>% summarize(newMeanSteps=mean(steps))
g <- ggplot(newMeanPerInterval, aes(x=interval, y=newMeanSteps)) + geom_line(color="blue")+ facet_grid(DayoftheWeek~.) + xlab("Interval") + ylab("Average Steps") + ggtitle("Average Steps - Comparison between Weekday and Weekend")
print(g)
```

```{r, echo=FALSE}
png("WeekdayAndWeekend.png")
dev.set(2)
dev.copy(png, "WeekdayAndWeekend.png")
dev.off()
```
